A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-13 paper/paper-monash/paper-monash.tex
monash preprint also

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{graphicx}

@@

\author{
    Dewi Amaliah
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dlamaleeah@gmail.com}{\nolinkurl{dlamaleeah@gmail.com}}} \\
   \And
    Dianne Cook
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
   \And
    Emi Tanaka
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}} \\
   \And
    Kate Hyde
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:hyde.kate.a@gmail.com}{\nolinkurl{hyde.kate.a@gmail.com}}} \\
   \And
    Nicholas Tierney
   \\
    Telethon Kids Institute \\
  Nedlands, WA 6009 \\
  \texttt{\href{mailto:nicholas.tierney@gmail.com}{\nolinkurl{nicholas.tierney@gmail.com}}} \\
  }

% Pandoc citation processing

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\def\tightlist{}

\begin{abstract}
Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to focus on methodology. Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection. It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today. This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s. The data is useful for teaching modeling and exploratory analysis of longitudinal data. Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles. The NLSY79 database has been continuously updated through to 2018, so new records are available. Here we describe our journey to refresh the wages data, and document the process so that the data can be regularly updated into the future. Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated. We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching. Three new data sets and the code to produce them are provided in the open source R package called \texttt{yowie}.
\end{abstract}

\keywords{
    Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;
  }

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average. However, an exploratory analysis reveals that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.

This disparity between the overall trend and the individual heterogeneity is what makes this data interesting. Textbook data sets have longevity if they have unresolved elements. The iris data (Anderson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)}. We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For statistics, and data science by association, it is increasingly important to reach the individual. One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous. Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics. Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak. The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable. However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different. It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility. It would be more useful for an individual to know what percentage of the population have profiles like theirs. We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation. We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015). Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.

As a textbook data set, though, the wages data is outdated. The most recent year in the data is 1994, 9 years prior to when Singer and Willett (2003) was published. Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data (see details at Open Knowledge Foundation 2021), which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open wild data to textbook data.

This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by Huebner, Vach, and Cessie (2016). They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Clean data often then goes through an ``initial data analysis'' (IDA) (Chatfield 1985), where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling. This stage is related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing. Dasu and Johnson (2003) say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011). Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998). There are a number of data sets provided by Singer and Willett (2003), but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.

@@

The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 (Bureau of Labor Statistics, U.S. Department of Labor 2021a). The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. For a variety of reasons, some structural, the number of respondents dropped to 9,964 after 1990. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

@@

There are two ways to conduct the interview of the NLSY79, which are face-to-face or telephone interviews. In recent survey years, more than 90 percent of respondents were interviewed by telephone (Cooksey 2017).

@@

The NLSY79 data used in Singer and Willett (2003) contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994. This dataset contains several variables as follows:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The wages data of the whole NLSY79 cohort, including females,
\item
  A separate table of the demographic data of the whole NLSY79 cohort, and
\item
  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of Singer and Willett (2003)'s data.
\end{enumerate}

@@

In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

To get the data into tidy form (Wickham 2014), it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns. The raw data consequently has a large number of columns (742 to be specific). The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data show the format problems:

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://numbats.github.io/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

>>>>

\documentclass[11pt,a4paper,]{article}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database},
            pdfkeywords={Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{../references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\RequirePackage{bera}
\RequirePackage{mathpazo}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*5}{*1}
\titlespacing{\subsection}{0pt}{*2}{*0.2}

%% TITLE PAGE
\def\Date{\number\day}
\def\Month{\ifcase\month\or
 January\or February\or March\or April\or May\or June\or
 July\or August\or September\or October\or November\or December\fi}
\def\Year{\number\year}

\makeatletter
\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
\def\nojel{\def\showjel{}}
\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
\def\cover{{\sffamily\setcounter{page}{0}
        \thispagestyle{empty}
        \placefig{2}{1.5}{width=5cm}{monash2}
        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
        \begin{textblock}{7}(12.7,27.9)\hfill
        \includegraphics[height=0.7cm]{AACSB}~~~
        \includegraphics[height=0.7cm]{EQUIS}~~~
        \includegraphics[height=0.7cm]{AMBA}
        \end{textblock}
        \vspace*{2cm}
        \begin{center}\Large
        Department of Econometrics and Business Statistics\\[.5cm]
        \footnotesize http://monash.edu/business/ebs/research/publications
        \end{center}\vspace{2cm}
        \begin{center}
        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
                \LARGE\@author\end{onehalfspace}
        }}
        \end{center}
        \vfill
                \begin{center}\Large
                \Month~\Year\\[1cm]
                Working Paper \@wp
        \end{center}\vspace*{2cm}}}
\def\pageone{{\sffamily\setstretch{1}%
        \thispagestyle{empty}%
        \vbox to \textheight{%
        \raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
        \vspace{2cm}\par
        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
        \hspace{1cm}\showjel\vss}}}
\def\blindtitle{{\sffamily
     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
        }}
\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}

\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
\let\maketitle\titlepage
\makeatother

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

%% KEYWORDS
\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}

%% ABSTRACT
\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
  {\end{minipage}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[showonlyrefs]{mathtools}
\usepackage[no-weekday]{eukdate}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

\wp{no/yr}
\jel{C10,C14,C22}

\RequirePackage[absolute,overlay]{textpos}
\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}
\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}

\author{}
\addresses{}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf : \@date}
\makeatother

%% Any special functions or other packages can be loaded here.
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

% Adjust headwidth in case user has changed geometry in header-includes
\renewcommand{\headwidth}{\textwidth}

\begin{document}
\maketitle
\begin{abstract}
Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to focus on methodology. Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection. It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today. This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s. The data is useful for teaching modeling and exploratory analysis of longitudinal data. Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles. The NLSY79 database has been continuously updated through to 2018, so new records are available. Here we describe our journey to refresh the wages data, and document the process so that the data can be regularly updated into the future. Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated. We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching. Three new data sets and the code to produce them are provided in the open source R package called \texttt{yowie}.
\end{abstract}
\begin{keywords}
Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;
\end{keywords}

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by \textcite{SingerJudithD2003Alda} in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) \autocite{nlsy79}.

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average. However, an exploratory analysis reveals that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in \textcite{ilk2004}, and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.

This disparity between the overall trend and the individual heterogeneity is what makes this data interesting. Textbook data sets have longevity if they have unresolved elements. The iris data \autocite{iris-data} is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data \autocite{penguins-data}, which has similar qualities, is new, and does not suffer from a connection with eugenics \autocite{notiris}}. We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For statistics, and data science by association, it is increasingly important to reach the individual. One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous. Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics. Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak. The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable. However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different. It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility. It would be more useful for an individual to know what percentage of the population have profiles like theirs. We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation. We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian \autocite{notaverage}. Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.

As a textbook data set, though, the wages data is outdated. The most recent year in the data is 1994, 9 years prior to when \textcite{SingerJudithD2003Alda} was published. Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data \autocite[see details at][]{opendata}, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open wild data to textbook data.

This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by \textcite{HuebnerMariannePhD2016Asat}. They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Clean data often then goes through an ``initial data analysis'' (IDA) \autocite{Chatfield1985TIEo}, where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling. This stage is related to exploratory data analysis (EDA), coined by \textcite{tukey} with a focus on learning from data. EDA can be considered to encompass IDA. In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing. \textcite{DasuTamraparni2003Edma} say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach \autocite{tidyverse}. The data is first organized into ``tidy data'' \autocite{WickhamHadley2014TD} and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables \autocite{plyr}. Tidy data shouldn't be confused with ``tame data,'' which \textcite{tamedata} coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

\textcite{SingerJudithD2003Alda} used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century \autocite{MichaelRPergamit2001DWTN}. In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research \autocite{MichaelRPergamit2001DWTN,eliznlsy}. According to \textcite{eliznlsy}, thousands of articles and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data \autocite{MichaelRPergamit2001DWTN}.

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with \textcite{grimshaw}'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by \textcite{SingerJudithD2003Alda}. We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} \autocite{racismnotrace}. There are a number of data sets provided by \textcite{SingerJudithD2003Alda}, but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.

@@

The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 \autocite{nlsy79}. The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. For a variety of reasons, some structural, the number of respondents dropped to 9,964 after 1990. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

@@

There are two ways to conduct the interview of the NLSY79, which are face-to-face or telephone interviews. In recent survey years, more than 90 percent of respondents were interviewed by telephone \autocite{eliznlsy}.

@@

The NLSY79 data used in \textcite{SingerJudithD2003Alda} contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994. This dataset contains several variables as follows:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The wages data of the whole NLSY79 cohort, including females,
\item
  A separate table of the demographic data of the whole NLSY79 cohort, and
\item
  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of \textcite{SingerJudithD2003Alda}'s data.
\end{enumerate}

@@

In the context of official statistics, \textcite{LooMarkvander2018Sdcw} describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

To get the data into tidy form \autocite{WickhamHadley2014TD}, it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns. The raw data consequently has a large number of columns (742 to be specific). The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data show the format problems:

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages \autocite{tidyverse}: \texttt{tidyr} \autocite{tidyr} to pivot the data into long-form, with \texttt{dplyr} \autocite{dplyr} and \texttt{stringr} \autocite{stringr} to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://numbats.github.io/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart}

@@

\printbibliography[title=References]
